---
title: "Analysis of periodontal bacteria in relation to disease status and pocket depth"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
The following notebook performs multidimensional scaling with the bacterial relative 
abundance data from PerioData.csv
The notebook follows the code from:
https://environmentalcomputing.net/graphics/multivariate-vis/mds/


```{r}
library(vegan)
library(ggplot2)
```


### Read Data

Make data frame from the periodontal data file PerioData.csv
```{r}
perio <- read.csv("PerioData.csv")
head(perio)
```
Subset the perio data to only have bacterial columns. 
The next block performs non-metric multidimentional scaling (NMDS) using bray-curtis distances of the bacterial relative abundances.
The plot shows all the samples ordinated on this first two axes, MDS1 and MDS2.

```{r}
bacteria <- perio[2:6]
#head(bacteria)
bacteria.mds <- metaMDS(comm = bacteria, distance = "bray", trace = FALSE, autotransform = FALSE)
plot(bacteria.mds$points)
```
This block assigns the metadata to the samples in the plot, then the ggplot colors the points in the graph for one variable, in this case disease status.

```{r}
bacteria_xy <- data.frame(bacteria.mds$points)
bacteria_xy$status <- perio$status #disease status 1=not disease; 2=diseased
bacteria_xy$time <- perio$time     #time point
bacteria_xy$range <- perio$range   #range of pocket depth shallow < 3mm, middle 3-3.9, deep 4+ 

ggplot(bacteria_xy, aes(MDS1, MDS2, color = status)) +
  geom_point() +
  theme_bw()
```
Finally, this block performs a PERMANOVA statistical analysis. This is an ANOVA using distance matrices that uses permutations to ask if the samples group clusters are the same or not. In effect, for this next test, we are asking if there is a difference between the beta diversity of the samples based on disease status.

Here is the full formula for the adnonis2 function that runs the PERMANOVA. By default it performs 999 permutation and uses bray curtis distances. 
adonis2(formula, data, permutations = 999, method = "bray",
    sqrt.dist = FALSE, add = FALSE, by = NULL,
    parallel = getOption("mc.cores"), na.action = na.fail,
    strata = NULL, ...)

```{r}
adonis2(bacteria ~ perio$status) # is there a significant different in the samples base on status?

```

There is no significant difference between the samples based upon the status because the p-value = 0.344. 

EXERCISE 1: Run the same plots and PERMANOVA using the other two variables: time and range

```{r} 
#looks at the time instead of status for the plot 
ggplot(bacteria_xy, aes(MDS1, MDS2, color = factor(time))) +
  geom_point(size = 3) +
  theme_bw() +
  labs(color = "Time Point") 

#looks at the range for the plot 
ggplot(bacteria_xy, aes(MDS1, MDS2, color = factor(range))) +
  geom_point(size = 3) +
  theme_bw() +
  labs(color = "Pocket Depth Range")
```

Conduct the PERMANOVA test 

```{r}
adonis2(bacteria ~ time, data = perio, method = "bray") #looks at time only  
adonis2(bacteria ~ range, data = perio, method = "bray") #looks at range only 
adonis2(bacteria ~ time + range, data = perio, method = "bray") # looks at both time + range 
```

Time alone does not impact the % of bacteria present (p = 0.138). The relationship with time + range does impact the % of bacteria present (p < 0.05), however since time alone does not impact the % of bacteria present, it could be heavily impacted by the range, which by itself does impact the % of bacteria present (p < 0.05). 

EXERCISE 2: REPEAT the analyses above after rclr transformation of the bacteria data.

Step 1) convert to rclr - zeros replaced with minimum values
Step 2) run the mds with method="euclidean" and plot with the same variables
Step 3) run the adonis2 with euclidean distances and 9999 permutations

Step 1: 

```{r}
bacteria <- perio[, c("strep", "prev", "lepto", "fuso", "veil")] #pull all the bacteria columns 
rclr_transform <- decostand(bacteria, method = "rclr") #robust clr transform  
column_names <- c("rclr_strep","rclr_prev","rclr_lepto",
                                     "rclr_fuso", "rclr_veil") #create the column names 
colnames(rclr_transform) <- column_names #rename the column names 
head(rclr_transform) #check the data  
perio <- cbind(perio,rclr_transform) #add rclr_transform to the perio data 
head(perio) #check the data 
```

Step 2: 

```{r}
rclr_bacteria <- perio[12:16] #select the rclr transformed data 

bacteria.mds <- metaMDS(comm = rclr_bacteria, distance = "euclidean", trace = FALSE, autotransform = FALSE)  #calculating the mds with euclidean distance + rclr_bacteria 


bacteria_xy <- data.frame(bacteria.mds$points)
bacteria_xy$status <- as.factor(perio$status) #disease status 1=not disease; 2=diseased
bacteria_xy$time <- perio$time #time point
bacteria_xy$range <- perio$range #range of pocket depth shallow < 3mm, middle 3-3.9, deep 4+

```

```{r}
#looks at the status 
ggplot(bacteria_xy, aes(x = MDS1, y = MDS2, color = status)) + #assign variables
  geom_point(size = 3) + #create scatter plot
  theme_bw() #denote theme 
```

```{r}
#looks at the time instead of status for the plot 
ggplot(bacteria_xy, aes(x = MDS1, y = MDS2, color = time)) + #assign variables
  geom_point(size = 3) + #create scatter plot
  theme_bw() #denote theme 

#looks at range instead of status for the plot 
ggplot(bacteria_xy, aes(x = MDS1, y = MDS2, color = range)) + #assign variables
  geom_point(size = 3) + #creates the scatter plot
  theme_bw() #denote the theme 
```


Step 3: 

```{r} 
#create a distance matrix for the PERMANOVA 
dist <- dist(rclr_bacteria, method = "euclidean") #use the rclr data 

#run the PERMANOVA utilizing euclidean distances and 9999 permutations
adonis2(dist ~ status, data = perio, permutations = 9999) #looking only at status  
adonis2(dist ~ time, data = perio, permutations = 9999) #looking only at time
adonis2(dist ~ range, data = perio, permutations = 9999) #looking only at range 
adonis2(dist ~ time + range, data = perio, permutations = 9999) #looking at time + range
```

