---
title: "R Project Code"
author: "Ariana Beloiu"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages Needed

```{r}
library(tidyverse) 
library(dplyr) 
library(car) 
library(ggfortify) 
library(vegan) 
library(ggplot2)
```

## Part A: Univariate statistics with R

### Import the data

```{r}
perio <- read.csv("PerioData.csv") #open the perio data  

head(perio) #inspect the data
```

### Make histograms, run qqnorm and qqline of all bacteria and “deepest”

```{r}
hist(perio$strep) #strep histogram 
qqnorm(perio$strep) #strep qq-plot  
qqline(perio$strep) #adds reference line to the qq-plot  

hist(perio$prev) #prev histogram 
qqnorm(perio$prev) #prev qq-plot  
qqline(perio$prev) #adds reference line to the qq-plot 

hist(perio$lepto) #lepto histogram 
qqnorm(perio$lepto) #lepto qq-plot  
qqline(perio$lepto) #adds reference line to the qq-plot  

hist(perio$fuso) #fuso histogram 
qqnorm(perio$fuso) #fuso qq-plot  
qqline(perio$fuso) #adds reference line to the qq-plot  

hist(perio$veil) #veil histogram 
qqnorm(perio$veil) #veil qq-plot  
qqline(perio$veil) #adds reference line to the qq-plot  

hist(perio$deepest) #"deepest" bacteria histogram 
qqnorm(perio$deepest) #"deepest" bacteria qq-plot  
qqline(perio$deepest) #adds reference line to the qq-plot 
```

The qqplots and histograms for each of the bacteria highlights that they are not following normality and display some skewness. This breaks the assumption of normality. To fix this we must transform the non-normal variables to better fit a normal distribution.

The deepest gum pocket data does approximately follow a normal distribution with minimal skewedness, so no transformation is required.

### Transform variables if non-normal and repeat histogram, qqnorm, qqline on transformed data. For transformations I suggest trying either sqrt() or log()

```{r}
tstrep = log(perio$strep) #log transform strep data  
hist(tstrep) #create a histogram for the transformed strep data 
qqnorm(tstrep) #create a qq-plot for the transformed strep data
qqline(tstrep) #add reference line to the qqplot  

tprev = sqrt(perio$prev) #square root transform prev data
hist(tprev) #create a histogram for the transformed prev data 
qqnorm(tprev) #create a qq-plot for the transformed prev data 
qqline(tprev) #add reference line to the qqplot  

tlepto = sqrt(perio$lepto) #square root transform lepto data 
hist(tlepto) #create a histogram for the transformed lepto data 
qqnorm(tlepto) #create a qq-plot for the transformed lepto data 
qqline(tlepto) #add reference line to the qqplot 

tfuso = sqrt(perio$fuso) #log transform fuso data 
hist(tfuso) #create a histogram for the transformed fuso data 
qqnorm(tfuso) #create a qq-plot for the transformed fuso data 
qqline(tfuso) #add a reference line to the qqplot 

tveil = log(perio$veil + 1) #log + 1 transform veil data 
hist(tveil) #create a histogram for the transformed veil data 
qqnorm(tveil) #create a qq-plot for the transformed veil data 
qqline(tveil) #add a reference line to the qqplot  

```

### One-way ANOVA and summary and boxplots:

```{r}
#Lepto by Time 

model1 <- aov(tlepto ~ time, data = perio) #create the ANOVA table 
model1 #print the anova table 
summary(model1) #print the summary table 
boxplot(tlepto ~ time, data = perio,
        xlab = "Time that the sample was taken",
        ylab = "Percentage of Leptotrichia bacteria", 
        names = c("Before gum cleaning", "After gum cleaning")) #create the boxplot 

```

After conducting a one-way ANOVA to test whether the percentage of *Leptotrichia* bacteria is affected by the time when gum cleaning occurred, it can be concluded that there were different levels of bacteria based on the time of when gum cleaning occurred (F(1,70), p-value \< 0.05).

```{r}
#Strep by time 

model2 <- aov(tstrep ~ time, data = perio) #create the ANOVA table 
model2 #print the anova table 
summary(model2) #print the summary table 
boxplot(tstrep ~ time, data = perio,
        xlab = "Time that the sample was taken",
        ylab = "Percentage of Streptococcus bacteria", 
        names = c("Before gum cleaning", "After gum cleaning")) #create the boxplot 

```

After conducting a one-way ANOVA to test whether the percentage of *Streptococcus* bacteria is affected by the time when gum cleaning occurred, it can be concluded that there was no significant difference between the levels of bacteria based on the time of when gum cleaning occurred (F(1,70), p-value = 0.354). 

### Correlation analysis (cor.test) 

```{r}

cor.test(tstrep, tprev) #conduct a correlation test between Strep and Prev Bacteria 
cor.test(tstrep, tfuso) #conduct a correlation test between Strep and Fuso Bacteria 
cor.test(tfuso, tlepto) #conduct a correlation test between Fuso and Lepto Bacteria 

```

After conducting correlation tests, it can be concluded that there is a correlation between the percentage of *Streptococcus* bacteria and the *Prevotella* bacteria (p-value < 0.05), as well as a correlation between the percentage *Streptococcus* bacteria and *Fusobacteria* bacteria (p-value < 0.05). However, it can be concluded that there is no correlation between the percentage of *Fusobacteria* bacteria and *Leptotrichia* bacteria (p-value = 0.4119).  

### Linear Regressions 

```{r} 
#Strep vs. deepest 
model3 <- lm(deepest ~ tstrep, data = perio) #create the linear model 
plot(tstrep, perio$deepest,
     xlab = "Log transformed % Streptococcus bacteria",
     ylab = "Depth of the deepest gum pocket in the mouth (mm)") #create the plot 
abline(model3, col = "blue") #adds a regression line to the plot 

```

```{r}
#Fuso vs. deepest 
model4 <- lm(deepest ~ tfuso, data = perio) #create the linear model 
plot(tfuso, perio$deepest,
     xlab = "Square root of the % Fusobacteria bacteria",
     ylab = "Depth of the deepest gum pocket in the mouth (mm)") #create the plot 
abline(model4, col = "blue") #adds a regression line to the plot 
```

After conducting linear regressions looking at *Fusobacteria*, *Streptococcus* bacteria, and their relationship to the depth of the deepest gum pocket in the mount (mm), we can see that their is a slightly negative relationship between Streptococcus bacteria in the mouth and the depth of the gum pocket in the mouth, indicating that as the greater the gum pocket, the lower the % of the *Streptococcus* bacteria in the mouth. We can also see a slightly positive relationship between the amount of *Fusobacteria* bacteria present in the mouth and the depth of the gum pocket in the mouth, indicating that as the amount of fusobacteria increases, the deeper the depth of the gum pocket in the mouth. 

## Part B: Use ggplot2 to make pretty graphs. 

### ggplot histogram of prev variable 

```{r}
prev_hist <- ggplot(data = perio, aes(x = prev)) + #assign the variable
  geom_histogram(binwidth = 5, color = "black", fill = "lightblue") + #create the histogram
  labs(x = "Prevotella bacteria",
       y = "Count") + #provide the labels 
  theme_classic() #change the theme 
prev_hist #print the plot
```

### ggplot scatterplot of strep by deepest, colored by time 

```{r} 
perio$time <- as.factor(perio$time) #convert time into a factor 
strep_deep_plot <- ggplot(data = perio, aes(x = strep, y = deepest, color = time)) + #assign the variables
  geom_point() + #create the scatter plot 
  labs(x = "% of Streptococcus Bacteria ", 
       y = "Depth of the deepest gum pocket in the mouth (mm)") +  #change the labels 
  theme_classic() #change the theme 
strep_deep_plot #print the plot
```

### ggplot scatterplot of fuso by deepest 

```{r}
fuso_deep_plot <- ggplot(data = perio, aes(x = fuso, y = deepest, color = time)) + #assign the variables
  geom_point() + #create the scatter plot 
  labs(x = "% of Fusobacteria Bacteria ", 
       y = "Depth of the deepest gum pocket in the mouth (mm)") +  #change the labels 
  theme_classic() #change the theme 
fuso_deep_plot #print the plot
```

### utilize the smoothing algorithm   
 
```{r}
prev_hist <- ggplot(data = perio, aes(x = prev)) + #assign the variable
  geom_histogram(binwidth = 5, color = "black", fill = "lightblue") + #create the histogram 
  geom_density(aes(y = after_stat(count)), color = "blue", linewidth = 1) + #adds the smoothing algorithm
  labs(x = "Prevotella bacteria",
       y = "Count") + #provide the labels 
  theme_classic() #change the theme 
prev_hist #print the plot
```

```{r} 
strep_deep_plot <- ggplot(data = perio, aes(x = strep, y = deepest, color = time)) + #assign the variables
  geom_point() + #create the scatter plot  
  geom_smooth() + #adds a regression line 
  labs(x = "% of Streptococcus Bacteria ", 
       y = "Depth of the deepest gum pocket in the mouth (mm)") +  #change the labels 
  theme_classic() #change the theme 
strep_deep_plot #print the plot
```

```{r}
fuso_deep_plot <- ggplot(data = perio, aes(x = fuso, y = deepest, color = time)) + #assign the variables
  geom_point() + #create the scatter plot 
  geom_smooth() + #adds a regression line
  labs(x = "% of Fusobacteria Bacteria ", 
       y = "Depth of the deepest gum pocket in the mouth (mm)") +  #change the labels 
  theme_classic() #change the theme 
fuso_deep_plot #print the plot
```


## Part C: rclr transform the variables (just the bacteria) and repeat all the statistical tests above.  

robust clr the data 

```{r}
bacteria <- perio[, c("strep", "prev", "lepto", "fuso", "veil")] #pull all the bacteria columns 
rclr_transform <- decostand(bacteria, method = "rclr") #robust clr transform  
column_names <- c("rclr_strep","rclr_prev","rclr_lepto",
                                     "rclr_fuso", "rclr_veil") #create the column names 
colnames(rclr_transform) <- column_names #rename the column names 
head(rclr_transform) #check the data 
```

```{r}
perio <- cbind(perio,rclr_transform) #add rclr_transform to the perio data 
head(perio) #check the data 
```


redo the statistical tests 

```{r} 
#Lepto vs time 

model4 <- aov(rclr_lepto ~ time, data = perio) #create the ANOVA table 
model4 #print the anova table 
summary(model4) #print the summary table  
boxplot(rclr_lepto ~ time, data = perio,
        xlab = "Time that the sample was taken",
        ylab = "RCLR transformed Lepto", 
        names = c("Before gum cleaning", "After gum cleaning")) #create the boxplot 

```

```{r}
#Strep by time 

model5 <- aov(rclr_strep ~ time, data = perio) #create the ANOVA table 
model5 #print the anova table 
summary(model5) #print the summary table 
boxplot(rclr_strep ~ time, data = perio,
        xlab = "Time that the sample was taken",
        ylab = "RCLR transformed Strep", 
        names = c("Before gum cleaning", "After gum cleaning")) #create the boxplot 
```

```{r}
cor.test(perio$rclr_strep, perio$rclr_prev) #conduct a correlation test between Strep and Prev Bacteria  
cor.test(perio$rclr_strep, perio$rclr_fuso) #conduct a correlation test between Strep and Fuso Bacteria
cor.test(perio$rclr_fuso, perio$rclr_lepto) #conduct a correlation test between Fuso and Lepto Bacteria 
```

```{r}
#Strep vs. deepest 
model6 <- lm(deepest ~ rclr_strep, data = perio) #create the linear model 
plot(perio$rclr_strep, perio$deepest,
     xlab = "Log transformed % Streptococcus bacteria",
     ylab = "Depth of the deepest gum pocket in the mouth (mm)") #create the plot 
abline(model6, col = "blue") #adds a regression line to the plot 

```

```{r}
#Fuso vs. deepest 
model7 <- lm(deepest ~ rclr_fuso, data = perio) #create the linear model 
plot(perio$rclr_fuso, perio$deepest,
     xlab = "Square root of the % Fusobacteria bacteria",
     ylab = "Depth of the deepest gum pocket in the mouth (mm)") #create the plot 
abline(model7, col = "blue") #adds a regression line to the plot 
```


