---
title: "R Notebook"
output: html_notebook
---
## RCLR Standardization, followed by imputation

```{r}
library(vegan)
```


### Read Data

Make data frame from toy_data.csv file
```{r}
df_toy_data <- read.csv("toy_data.csv", row.names = 1)
head(df_toy_data)
```
### RCLR transformation w/o imputation

Use the non-vegan decostand with impute set to FALSE
All the zeros will be replaced with NA
```{r}
df_toy_data.rclr <- decostand(df_toy_data, "rclr", impute = FALSE)
head(df_toy_data.rclr)
```
### Replace NAs

Replace the NAs with zeros (prior to minimum value substitution)
```{r}
##Replace NA with 0
df_toy_data.rclr[is.na(df_toy_data.rclr)] <- 0
head(df_toy_data.rclr)
```

### Determine min value

Find the minimum value of the dataset:
```{r}
toy_data_rclr_min_value <- min(df_toy_data.rclr)
toy_data_rclr_min_value
```

## Replace 0 with min value

```{r}
df_toy_data_rclr_min_sub = df_toy_data.rclr # Copy the data

df_toy_data_rclr_min_sub[df_toy_data_rclr_min_sub == 0] = toy_data_rclr_min_value
head(df_toy_data_rclr_min_sub)

```



